---
title: "Portfolio Risk Analytics"
subtitle: "Calendar-consistent forecasts and evidence against benchmark models"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
table_path <- function(name) file.path("..", "outputs", "tables", paste0(name, ".csv"))
manifest_path <- table_path("output_manifest")
if (!file.exists(manifest_path)) stop("Run the analysis before rendering this report.")
manifest <- read.csv(manifest_path, check.names = FALSE)
if (!all(c("path", "md5", "run_id") %in% names(manifest))) {
  stop("The output manifest predates the current report schema; rerun the analysis.")
}
artifact_paths <- file.path("..", manifest$path)
if (any(!file.exists(artifact_paths))) stop("A manifested output is missing; rerun the analysis.")
actual_hashes <- unname(tools::md5sum(artifact_paths))
if (anyNA(manifest$md5) || any(actual_hashes != manifest$md5)) {
  stop("Output hashes do not match the run manifest; refusing to mix analysis runs.")
}
read_output <- function(name, required = TRUE) {
  relative <- paste0("outputs/tables/", name, ".csv")
  if (!relative %in% manifest$path) {
    if (required) stop(sprintf("Required table is absent from this run: %s", name))
    return(NULL)
  }
  read.csv(table_path(name), check.names = FALSE)
}
show_columns <- function(x, columns, digits = 5) {
  knitr::kable(x[, intersect(columns, names(x)), drop = FALSE], digits = digits, row.names = FALSE)
}
figure_path <- function(name) file.path("..", "outputs", "figures", name)
show_figure <- function(name) {
  if (paste0("outputs/figures/", name) %in% manifest$path) {
    knitr::include_graphics(figure_path(name))
  }
}
provenance <- read_output("run_provenance")
provenance_value <- function(key) {
  value <- provenance$value[match(key, provenance$key)]
  if (!length(value) || is.na(value)) "unavailable" else value
}
if (length(unique(manifest$run_id)) != 1L ||
    unique(manifest$run_id) != provenance_value("run_id")) {
  stop("Provenance and output manifest identify different runs.")
}
rolling <- read_output("rolling_forecasts")
comparison <- read_output("model_comparison")
if (!all(c("loss_var", "loss_es") %in% names(rolling)) ||
    !all(c("common_observations", "mean_joint_score") %in% names(comparison))) {
  stop("Forecast outputs use an obsolete risk or evaluation schema; rerun the analysis.")
}
forecast_dates <- sort(unique(as.Date(rolling$forecast_date)))
oos_start <- as.character(min(forecast_dates))
oos_end <- as.character(max(forecast_dates))
oos_count <- length(forecast_dates)
audit <- read_output("valuation_audit")
```

# Research question and observed scope

Does conditional volatility improve portfolio risk forecasts over simple benchmarks,
and does modelling cross-asset dependence add value once the marginal forecasts are held
fixed? The tables below answer these as separate comparisons.

This report uses run **`r provenance_value("run_id")`**, profile
**`r provenance_value("profile")`**, and source **`r provenance_value("data_source")`**.
The input contains **`r provenance_value("observations")`** valuation observations from
**`r provenance_value("valuation_start")`** to **`r provenance_value("valuation_end")`**.
The actual out-of-sample evaluation runs from **`r oos_start`** to **`r oos_end`** across
**`r oos_count` distinct forecast dates** and **`r length(unique(rolling$model))` models**.
Input history and evaluation history are different samples.

```{r}
#| results: asis
if (provenance_value("data_source") == "synthetic_fixture") {
  cat("**This run uses synthetic fixture data. It verifies software behaviour and does not establish real-market forecast performance.**\n")
}
```

The current estimate below is conditional on the last available valuation date; it is
separate from the historical forecasts used to assess calibration and predictive loss.
No model is declared superior from an in-sample fit, a coverage p-value, or a point rank.

```{r}
show_columns(read_output("current_copula_risk"),
  c("confidence", "loss_var", "loss_es", "var", "es", "tail_effective_observations"))
```

# Valuation calendar and portfolio definition

Weekly profiles use a shared Friday valuation calendar. For each asset, the input is
its most recent quote dated on or before that Friday, within the permitted age limit.
The quote date remains visible below: a Thursday market close can be used for a Friday
holiday valuation without deleting that week. Future quotes, an entirely missing week,
and excessive staleness are rejected. An unfinished final calendar week is excluded.
Daily profiles use common observed trading dates.

```{r}
calendar_summary <- do.call(rbind, lapply(split(audit, audit$ticker), function(x) {
  data.frame(asset = x$ticker[[1]], valuations = nrow(x),
             first_valuation = min(as.Date(x$valuation_date)),
             last_valuation = max(as.Date(x$valuation_date)),
             stale_quotes = sum(x$stale_days > 0), maximum_age_days = max(x$stale_days))
}))
knitr::kable(calendar_summary, row.names = FALSE)
```

```{r}
show_figure("asset_prices.png")
```

The configured return space is **`r provenance_value("return_type")`** and the currency
mode is **`r provenance_value("currency_mode")`**. Fixed weights apply to asset simple
returns at each observation. Portfolio log return is `log1p(weighted_simple_return)`.
Local-index-return mode combines abstract index return series and includes no FX
conversion; it is not an investable base-currency portfolio. Maintaining fixed weights
also implies rebalancing, whose costs are outside the current demonstration.

For realised return $r$, signed loss is $L=-r$. The `loss_var` and `loss_es` columns retain
the signed forecasts used for scoring, including negative risk estimates when the tail
still earns a gain. The `var` and `es` columns are nonnegative display magnitudes only.
VaR is a loss-quantile threshold, not a maximum loss. Empirical ES averages exactly the
worst tail probability mass, with a fractional observation at the boundary when needed.

# Models and diagnostics

Benchmarks include historical, Gaussian, Student-t, EWMA and filtered-historical risk,
and, when configured, a fixed portfolio sGARCH(1,1) model with Student-t innovations
(`garch_t`). The last benchmark tests whether a conditional volatility model fitted
directly to portfolio returns can account for the apparent benefit of the larger system.
The current benchmark estimates are shown separately from their out-of-sample scores.

```{r}
show_columns(read_output("current_baseline_risk"),
  c("model", "confidence", "loss_var", "loss_es", "status"))
```

Marginal selection screens convergence, parameter validity, and residual and squared
residual autocorrelation before information-criterion ranking. A selected fallback that
fails diagnostics remains visibly flagged. PIT goodness-of-fit p-values use fitted
residuals and are exploratory; they do not provide calibrated uniformity tests and are
not hard model-admission thresholds.

```{r}
selection <- read_output("marginal_model_selection")
selected <- selection[selection$selected %in% TRUE, , drop = FALSE]
show_columns(selected, c("asset", "model_id", "engine", "bic", "diagnostics_passed",
                         "rejection_reason", "pit_ks_p"), digits = 4)
```

```{r}
selection_counts <- do.call(rbind, lapply(split(selection, selection$asset), function(x) {
  data.frame(asset = x$asset[[1]], candidates = nrow(x),
             converged = sum(x$convergence %in% TRUE),
             diagnostics_passed = sum(x$diagnostics_passed %in% TRUE),
             selected_fallback = any(x$selected %in% TRUE & !x$diagnostics_passed %in% TRUE))
}))
knitr::kable(selection_counts, row.names = FALSE)
```

```{r}
show_columns(read_output("copula_model_selection"),
  c("family", "engine", "status", "bic", "kendall_tau", "lower_tail", "upper_tail", "selected"), digits = 4)
show_figure("copula_diagnostic.png")
```

Copula coefficients use the return-CDF scale: lower-tail dependence concerns joint
extreme losses, whereas upper-tail dependence concerns joint extreme gains. These
coefficients describe the fitted model, and do not establish its predictive value.

# Out-of-sample calibration and availability

Each forecast is fitted using observations through its training end and evaluated on
the next valuation. Models are refitted at each origin; specification selection follows
the configured schedule. Failures remain on the evaluation calendar, so independence
tests do not join observations across a missing forecast.

```{r}
calibration <- comparison
calibration$expected_exceptions <- calibration$observations * (1 - calibration$confidence)
show_columns(calibration, c("model", "confidence", "observations", "availability",
                            "model_failures", "exceptions", "expected_exceptions", "coverage_status"), digits = 3)
```

```{r}
show_columns(comparison, c("model", "confidence", "kupiec_p_value", "independence_p_value",
                          "independence_status", "transition_pairs", "conditional_coverage_p_value"), digits = 4)
```

Expected exception counts show the information available for tail calibration. With
few expected or observed exceptions, a large p-value has limited evidential value;
failure to reject is not proof of an adequate model. Independence can be unidentified
at boundary cases. The reported statuses and transition counts must be read alongside
p-values. Descriptive ES exceedance residuals are not a formal ES hypothesis test.

```{r}
show_figure("rolling_var_exceedances.png")
```

# Predictive loss and dependence ablations

Quantile-loss ranks use dates for which every compared model has a valid VaR forecast.
Joint VaR/ES ranks use a separate shared set of valid FZ0 scores. Their sample counts
are displayed explicitly; model-specific losses are not silently ranked on different
dates. Runtime and complexity remain operational information and are not mixed into
a hand-weighted statistical score.

```{r}
show_columns(comparison, c("model", "confidence", "common_observations", "quantile_loss", "rank",
                          "joint_common_observations", "mean_joint_score", "joint_score_rank"), digits = 7)
```

FZ0 jointly scores signed VaR and ES and requires strictly positive signed ES. Otherwise
valid forecasts outside this score's domain are unavailable for FZ0, rather than clipped
or assigned an artificial score. Lower mean scores are preferable on the same sample
and at the same confidence level; negative FZ0 values are possible.

```{r}
show_columns(comparison, c("model", "confidence", "joint_score_availability", "joint_score_status",
                          "tail_observations", "mean_exceedance_residual"), digits = 5)
```

When configured, `copula_independence`, `copula_gaussian`, `copula_student` and
`copula_bb1` reuse the **same conditional marginal forecasts** at each origin while
changing the dependence family. `copula_garch` uses the selection procedure. Differences
within this group address the contribution of dependence; differences against
`garch_t` address the value of the full multivariate construction. A failed fixed-family
fit is recorded as a failure, rather than substituted with another family.

```{r}
show_figure("model_comparison.png")
```

# Uncertainty in paired score differences

```{r}
predictive <- read_output("predictive_comparison")
show_columns(predictive, c("model", "reference_model", "confidence", "score_type",
                          "observations", "mean_loss_difference", "ci_lower", "ci_upper"), digits = 7)
```

Differences are **model minus reference**: negative values favour the model. The circular
block bootstrap resamples paired score differences to retain local serial dependence.
It reports percentile intervals and centred two-sided bootstrap p-values, with Holm
adjustment across the comparisons in this run. The intervals themselves are marginal,
not simultaneous. An interval containing zero does not resolve the direction of the
difference, and a point rank does not establish superiority.

```{r}
#| results: asis
main_pairs <- predictive[predictive$model == "copula_garch", , drop = FALSE]
if (nrow(main_pairs)) {
  for (i in seq_len(nrow(main_pairs))) {
    pair <- main_pairs[i, ]
    metric <- if (pair$score_type == "quantile_loss") "quantile loss" else "joint VaR/ES score"
    label <- sprintf("At %.1f%% confidence, copula-GARCH versus **%s** has mean %s difference **%s**",
                     100 * pair$confidence, pair$reference_model, metric,
                     format(pair$mean_loss_difference, digits = 4, scientific = TRUE))
    if (pair$status != "ok") {
      cat(sprintf("%s. Paired inference is unavailable: `%s`.\n\n", label, pair$status))
    } else {
      direction <- if (pair$ci_upper < 0) "The marginal interval lies below zero." else if (pair$ci_lower > 0) {
        "The marginal interval lies above zero."
      } else "The interval crosses zero, so its direction is unresolved."
      cat(sprintf("%s, with a %.1f%% interval [%s, %s] and Holm-adjusted p-value %s. %s\n\n",
                  label, 100 * pair$confidence_interval,
                  format(pair$ci_lower, digits = 4, scientific = TRUE),
                  format(pair$ci_upper, digits = 4, scientific = TRUE),
                  format(pair$p_value_holm, digits = 4), direction))
    }
  }
} else {
  cat("The main copula-GARCH model is absent from this run's paired comparisons; this run does not evaluate its incremental predictive value.\n")
}
```

```{r}
show_columns(predictive, c("model", "reference_model", "confidence", "score_type",
                          "common_sample_fraction", "block_length", "bootstrap_replications",
                          "p_value_holm", "status"), digits = 5)
```

Comparisons use the same common date set across all models at a confidence level and
score type. If that common set has internal calendar gaps, the mean difference remains
visible but bootstrap inference is unavailable, rather than bridging those gaps.
Insufficient samples also have an explicit unavailable status. Per-date paired scores
and differences are retained in `predictive_scores.csv`.

# Monte Carlo precision

```{r}
convergence <- read_output("monte_carlo_convergence_summary", required = FALSE)
if (!is.null(convergence)) {
  show_columns(convergence, c("n_simulations", "confidence", "repetitions", "mean_var", "sd_var",
                              "mean_es", "sd_es"), digits = 6)
} else {
  knitr::kable(data.frame(Status = "The convergence experiment was not enabled for this run."))
}
```

```{r}
show_figure("monte_carlo_convergence.png")
```

Across-seed variation measures simulation error conditional on the fitted model. It
does not include uncertainty from estimated parameters, model selection, historical
sample choice, or structural change. Simulation counts, seed and chunk size are recorded
below; identical seeds alone do not guarantee identical results after changing chunk size.

# Runtime, provenance and limitations

```{r}
show_columns(comparison, c("model", "confidence", "runtime_seconds", "model_failures"), digits = 3)
```

Runtime is elapsed wall-clock time, not CPU time. Copula ablations reuse marginal fits;
their model-attributed times include that shared marginal cost, so summing model rows
does not measure the runtime of the whole pipeline. Confidence-level rows may also
share a fit or simulation.

```{r}
keys <- c("run_id", "profile", "git_commit", "source_dirty", "implementation_hash", "config_hash",
          "input_hash", "aligned_hash", "retrieved_at_utc", "r_version", "rng_kind",
          "simulation_seed", "current_draws", "rolling_draws", "chunk_size")
knitr::kable(provenance[match(keys, provenance$key, nomatch = 0L), ], row.names = FALSE)
```

The run archive retains its configuration, price snapshot and session information.
The manifest records output hashes; this report checks them before reading the results.
Package versions are listed in `run_provenance.csv`. A dependency lockfile fixes software
versions, while the archived price snapshot fixes the market inputs for a repeat run.

This remains a bivariate research system. Local-index-return results omit FX and
implementation costs. One observed historical evaluation period cannot establish
performance in every market regime. Calendar correctness, common-sample scoring and
recorded provenance make the experiment auditable; predictive superiority still requires
adequate samples and stable evidence across independent comparisons.
