---
title: "Volatility and Dependence in Portfolio Tail-Risk Forecasting"
subtitle: "A rolling out-of-sample study of copula-GARCH VaR and Expected Shortfall"
bibliography: references.bib
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
table_path <- function(name) file.path("..", "outputs", "tables", paste0(name, ".csv"))
manifest_path <- table_path("output_manifest")
if (!file.exists(manifest_path)) stop("Run the analysis before rendering this report.")
manifest <- read.csv(manifest_path, check.names = FALSE)
if (!all(c("path", "md5", "run_id") %in% names(manifest))) {
  stop("The output manifest predates the current report schema; rerun the analysis.")
}
artifact_paths <- file.path("..", manifest$path)
if (any(!file.exists(artifact_paths))) stop("A manifested output is missing; rerun the analysis.")
actual_hashes <- unname(tools::md5sum(artifact_paths))
if (anyNA(manifest$md5) || any(actual_hashes != manifest$md5)) {
  stop("Output hashes do not match the run manifest; refusing to mix analysis runs.")
}
read_output <- function(name, required = TRUE) {
  relative <- paste0("outputs/tables/", name, ".csv")
  if (!relative %in% manifest$path) {
    if (required) stop(sprintf("Required table is absent from this run: %s", name))
    return(NULL)
  }
  read.csv(table_path(name), check.names = FALSE)
}
show_columns <- function(x, columns, digits = 5) {
  knitr::kable(x[, intersect(columns, names(x)), drop = FALSE], digits = digits, row.names = FALSE)
}
figure_path <- function(name) file.path("..", "outputs", "figures", name)
show_figure <- function(name) {
  if (paste0("outputs/figures/", name) %in% manifest$path) {
    knitr::include_graphics(figure_path(name))
  }
}
provenance <- read_output("run_provenance")
provenance_value <- function(key) {
  value <- provenance$value[match(key, provenance$key)]
  if (!length(value) || is.na(value)) "unavailable" else value
}
if (length(unique(manifest$run_id)) != 1L ||
    unique(manifest$run_id) != provenance_value("run_id")) {
  stop("Provenance and output manifest identify different runs.")
}
rolling <- read_output("rolling_forecasts")
comparison <- read_output("model_comparison")
if (!all(c("loss_var", "loss_es") %in% names(rolling)) ||
    !all(c("common_observations", "mean_joint_score") %in% names(comparison))) {
  stop("Forecast outputs use an obsolete risk or evaluation schema; rerun the analysis.")
}
forecast_dates <- sort(unique(as.Date(rolling$forecast_date)))
oos_start <- as.character(min(forecast_dates))
oos_end <- as.character(max(forecast_dates))
oos_count <- length(forecast_dates)
audit <- read_output("valuation_audit")
```

# Research question and empirical design

Portfolio tail risk depends on both the evolution of individual asset volatility and
the tendency of assets to experience losses together. This study evaluates their
separate contributions to one-step Value-at-Risk (VaR) and Expected Shortfall (ES)
forecasting. A univariate portfolio GARCH-t benchmark captures conditional volatility
and heavy tails; copula models fitted to a common set of conditional marginals isolate
the additional contribution of dependence.

The sample from **`r provenance_value("data_source")`** contains
**`r provenance_value("observations")`** valuation observations from
**`r provenance_value("valuation_start")`** to **`r provenance_value("valuation_end")`**.
The out-of-sample experiment covers **`r oos_start`** through **`r oos_end`**, with
**`r oos_count` forecast dates** and **`r length(unique(rolling$model))` models**.
The evaluation combines coverage tests, common-sample predictive losses, and paired
uncertainty estimates. This distinguishes the size of an observed forecasting gain
from the strength of the evidence supporting it.

```{r}
#| results: asis
if (provenance_value("data_source") == "synthetic_fixture") {
  cat("**This run uses synthetic fixture data. It verifies software behaviour and does not establish real-market forecast performance.**\n")
}
```

The table below gives the conditional risk estimate at the final valuation date.
Historical rolling forecasts provide the evidence on calibration and predictive
performance developed in the subsequent sections.

```{r}
show_columns(read_output("current_copula_risk"),
  c("confidence", "loss_var", "loss_es", "var", "es", "tail_effective_observations"))
```

# Data and portfolio construction

Weekly observations are formed on a shared Friday valuation calendar. Each asset is
valued using its most recent quote available by the cutoff, subject to the configured
quote-age limit. This convention accommodates differences in market holidays while
maintaining a seven-day observation interval. The final sample includes complete
calendar weeks with an admissible quote for every asset. The table records sample
coverage and the age of the underlying quotes. Daily profiles use common observed
trading dates.

```{r}
calendar_summary <- do.call(rbind, lapply(split(audit, audit$ticker), function(x) {
  data.frame(asset = x$ticker[[1]], valuations = nrow(x),
             first_valuation = min(as.Date(x$valuation_date)),
             last_valuation = max(as.Date(x$valuation_date)),
             stale_quotes = sum(x$stale_days > 0), maximum_age_days = max(x$stale_days))
}))
knitr::kable(calendar_summary, row.names = FALSE)
```

```{r}
show_figure("asset_prices.png")
```

The return convention is **`r provenance_value("return_type")`** and the currency mode
is **`r provenance_value("currency_mode")`**. Asset simple returns are combined using
fixed portfolio weights. Portfolio log return is `log1p(weighted_simple_return)`.
Under the local-index-return convention, the portfolio combines index returns in their
respective currencies. The experiment therefore measures joint index-return risk;
FX exposure and the costs of maintaining fixed weights are outside its scope.

For realised return $r$, signed loss is $L=-r$. The `loss_var` and `loss_es` columns retain
the signed forecasts used for scoring, including negative risk estimates when the tail
still earns a gain. The `var` and `es` columns are nonnegative display magnitudes only.
VaR is a confidence-level loss quantile. Empirical ES integrates the worst tail
probability mass, with a fractional observation at the boundary when needed.

# Volatility and dependence models

The benchmark set includes historical, Gaussian, Student-t, EWMA and
filtered-historical forecasts. The portfolio sGARCH(1,1) model with Student-t
innovations (`garch_t`), when included, provides a direct conditional-volatility
alternative to the multivariate specification. Current estimates for the benchmark
models are reported below.

```{r}
show_columns(read_output("current_baseline_risk"),
  c("model", "confidence", "loss_var", "loss_es", "status"))
```

For asset $i$, the conditional marginal model has the form
$r_{i,t}=\mu_{i,t}+\sigma_{i,t}z_{i,t}$, with an ARMA conditional mean and a
GARCH-family conditional variance. Candidate innovation distributions allow for
heavy tails and asymmetry. Probability integral transforms of the standardized
innovations provide the inputs to the dependence model.

Marginal specification selection combines convergence and parameter checks,
residual and squared-residual autocorrelation diagnostics, and information-criterion
ranking. The selected specifications and screening results are reported below.
Fallback selections carry their diagnostic status. PIT goodness-of-fit statistics
provide descriptive checks on fitted residuals; their conventional p-values are
exploratory because the marginal parameters are estimated from the same observations.

```{r}
selection <- read_output("marginal_model_selection")
selected <- selection[selection$selected %in% TRUE, , drop = FALSE]
show_columns(selected, c("asset", "model_id", "engine", "bic", "diagnostics_passed",
                         "rejection_reason", "pit_ks_p"), digits = 4)
```

```{r}
selection_counts <- do.call(rbind, lapply(split(selection, selection$asset), function(x) {
  data.frame(asset = x$asset[[1]], candidates = nrow(x),
             converged = sum(x$convergence %in% TRUE),
             diagnostics_passed = sum(x$diagnostics_passed %in% TRUE),
             selected_fallback = any(x$selected %in% TRUE & !x$diagnostics_passed %in% TRUE))
}))
knitr::kable(selection_counts, row.names = FALSE)
```

```{r}
show_columns(read_output("copula_model_selection"),
  c("family", "engine", "status", "bic", "kendall_tau", "lower_tail", "upper_tail", "selected"), digits = 4)
show_figure("copula_diagnostic.png")
```

The copula links the conditional marginal distributions to form a joint return
forecast, providing a framework for asymmetric dependence [@patton2006]. On the
return-CDF scale, lower-tail dependence characterizes joint extreme
losses and upper-tail dependence characterizes joint extreme gains. Joint simulated
returns are aggregated in simple-return space and converted to the portfolio return
convention before estimating VaR and ES. The fitted dependence coefficients describe
the current joint distribution; their forecasting contribution is assessed out of sample.

# Out-of-sample calibration

At each origin, estimation uses observations available through the training end and
forecasts the next valuation. Parameters are refitted at every origin, while
specification selection follows the configured schedule. Forecast availability is
reported alongside the statistical results.

Unconditional coverage evaluates whether the frequency of VaR exceptions matches
the nominal tail probability [@kupiec1995]. Independence and conditional coverage
also examine the sequence of exceptions [@christoffersen1998]. Transition counts
use adjacent available forecasts on the evaluation calendar.

```{r}
calibration <- comparison
calibration$expected_exceptions <- calibration$observations * (1 - calibration$confidence)
show_columns(calibration, c("model", "confidence", "observations", "availability",
                            "model_failures", "exceptions", "expected_exceptions", "coverage_status"), digits = 3)
```

```{r}
show_columns(comparison, c("model", "confidence", "kupiec_p_value", "independence_p_value",
                          "independence_status", "transition_pairs", "conditional_coverage_p_value"), digits = 4)
```

Expected exception counts indicate the effective information available for tail
calibration. Coverage results are interpreted together with these counts, transition
availability, and predictive losses. Small tail samples limit test power, and boundary
cases can leave independence unidentified. ES exceedance residuals provide a
complementary descriptive measure of tail forecast error.

```{r}
show_figure("rolling_var_exceedances.png")
```

# Predictive performance and dependence comparisons

Mean quantile loss evaluates the accuracy of VaR forecasts. Mean FZ0 evaluates VaR
and ES jointly. Each score uses the dates shared by all eligible models at a given
confidence level, with separate sample counts for the two scoring rules. Lower mean
loss indicates better observed predictive performance, and ranks summarize that
ordering. Estimation cost is considered separately.

```{r}
show_columns(comparison, c("model", "confidence", "common_observations", "quantile_loss", "rank",
                          "joint_common_observations", "mean_joint_score", "joint_score_rank"), digits = 7)
```

FZ0 is defined for strictly positive signed ES and permits negative values of the
score itself. Joint VaR/ES scoring provides the evaluation framework described by
@pattonziegelchen2019. Forecasts outside this domain remain part of the availability accounting
but do not enter the FZ0 comparison. Score levels are compared within the same
confidence level and observation sample.

```{r}
show_columns(comparison, c("model", "confidence", "joint_score_availability", "joint_score_status",
                          "tail_observations", "mean_exceedance_residual"), digits = 5)
```

The dependence experiment holds the **conditional marginal forecasts fixed** at each
origin and compares independence, Gaussian, Student-t and BB1 copulas. These appear
as `copula_independence`, `copula_gaussian`, `copula_student` and `copula_bb1`.
The `copula_garch` forecast applies the configured copula-selection procedure.
Comparisons within this group isolate dependence assumptions, while comparison with
`garch_t` measures the incremental value of the multivariate construction. Forecast
availability includes any unsuccessful fixed-family fits.

```{r}
show_figure("model_comparison.png")
```

# Paired evidence on incremental predictive value

```{r}
predictive <- read_output("predictive_comparison")
show_columns(predictive, c("model", "reference_model", "confidence", "score_type",
                          "observations", "mean_loss_difference", "ci_lower", "ci_upper"), digits = 7)
```

```{r}
#| label: fig-paired-uncertainty
#| fig-width: 12
#| fig-height: 9
#| fig-cap: "Mean score differences relative to the reference model. Negative values favour the model. Bars show the reported marginal bootstrap intervals (95% by default), not simultaneous intervals. Open points have no available interval; they retain the observed mean only. Interval colours do not encode Holm-adjusted p-values."
plot_paired_uncertainty <- function(x) {
  previous <- par(no.readonly = TRUE)
  on.exit(par(previous), add = TRUE)
  if (!nrow(x)) {
    plot.new()
    text(0.5, 0.55, "No paired predictive comparisons", cex = 1.2)
    text(0.5, 0.45, "No uncertainty interval is available for this run.", col = "#52616B")
    return(invisible(NULL))
  }
  metrics <- c("quantile_loss", "joint_var_es")
  levels <- sort(unique(x$confidence[is.finite(x$confidence)]))
  if (!length(levels)) {
    plot.new()
    text(0.5, 0.5, "No valid confidence levels in the paired comparisons.")
    return(invisible(NULL))
  }
  par(mfrow = c(length(metrics), length(levels)), mar = c(4.5, 10, 4.2, 1.2),
      oma = c(2.5, 0, 1, 0), mgp = c(2.8, 0.7, 0), tcl = -0.25)
  models <- sort(unique(as.character(x$model)))
  labels <- gsub("_", " ", models, fixed = TRUE)
  interval_ok <- function(z) z$status == "ok" & is.finite(z$ci_lower) &
    is.finite(z$ci_upper) & is.finite(z$mean_loss_difference)
  for (metric in metrics) {
    metric_rows <- x[x$score_type == metric, , drop = FALSE]
    valid_intervals <- interval_ok(metric_rows)
    values <- c(0, metric_rows$mean_loss_difference,
                metric_rows$ci_lower[valid_intervals], metric_rows$ci_upper[valid_intervals])
    limits <- range(values[is.finite(values)])
    width <- diff(limits)
    if (!is.finite(width) || width == 0) width <- max(abs(limits), 1e-6)
    limits <- limits + c(-1, 1) * 0.08 * width
    metric_label <- if (metric == "quantile_loss") "Quantile loss" else "Joint VaR/ES score"
    for (level in levels) {
      panel <- metric_rows[metric_rows$confidence == level, , drop = FALSE]
      title_text <- sprintf("%s | %.1f%% forecast confidence", metric_label, 100 * level)
      if (!nrow(panel) || !any(is.finite(panel$mean_loss_difference))) {
        plot.new()
        title(main = title_text, cex.main = 0.9)
        text(0.5, 0.55, "No eligible paired score observations", cex = 0.9)
        text(0.5, 0.43, "Inference is unavailable.", cex = 0.85, col = "#52616B")
        next
      }
      y <- length(models) + 1L - match(panel$model, models)
      plot(NA_real_, NA_real_, xlim = limits, ylim = c(0.4, length(models) + 0.6),
           xlab = "Mean score difference (model - reference)", ylab = "", axes = FALSE)
      abline(h = seq_along(models), col = "#EDF0F2", lwd = 0.8)
      abline(v = 0, col = "#73808C", lty = 2, lwd = 1)
      ticks <- pretty(limits, n = 4)
      axis(1, at = ticks, labels = format(ticks, scientific = TRUE, digits = 2, trim = TRUE), cex.axis = 0.8)
      axis(2, at = rev(seq_along(models)), labels = labels, las = 1, tick = FALSE, cex.axis = 0.85)
      box(col = "#C6CED3")
      available <- interval_ok(panel)
      if (any(available)) {
        segments(panel$ci_lower[available], y[available], panel$ci_upper[available], y[available],
                 col = "#244A64", lwd = 1.7)
        segments(panel$ci_lower[available], y[available] - 0.12,
                 panel$ci_lower[available], y[available] + 0.12, col = "#244A64")
        segments(panel$ci_upper[available], y[available] - 0.12,
                 panel$ci_upper[available], y[available] + 0.12, col = "#244A64")
      }
      finite <- is.finite(panel$mean_loss_difference)
      points(panel$mean_loss_difference[finite], y[finite], pch = ifelse(available[finite], 19, 1),
             col = "#244A64", cex = 0.8)
      if (any(!finite)) text(limits[[2]], y[!finite], "unavailable", adj = 1,
                            cex = 0.75, col = "#73808C")
      title(main = title_text, cex.main = 0.9)
      interval_levels <- unique(panel$confidence_interval[available])
      interval_label <- if (length(interval_levels) == 1L) {
        sprintf("%.1f%% marginal intervals", 100 * interval_levels)
      } else if (length(interval_levels)) "Reported marginal intervals" else "Intervals unavailable"
      mtext(sprintf("Reference: %s | %s", paste(unique(panel$reference_model), collapse = ", "),
                    interval_label), side = 3, line = 0.35, cex = 0.7, col = "#52616B")
    }
  }
  mtext("Lower differences favour the model; zero means equal observed mean loss.",
        side = 1, outer = TRUE, line = 1, cex = 0.9)
  invisible(NULL)
}
plot_paired_uncertainty(predictive)
```

The paired mean difference measures the observed forecasting gain relative to the
reference model. Differences are **model minus reference**, so negative values favour
the model. A circular block bootstrap retains local serial dependence in the paired
scores and estimates uncertainty around their mean difference. The analysis reports
marginal percentile intervals and centred two-sided bootstrap p-values, with Holm
adjustment across the comparisons. The intervals are interpreted individually;
the adjusted p-values address multiplicity across the experiment.

```{r}
#| results: asis
main_pairs <- predictive[predictive$model == "copula_garch", , drop = FALSE]
if (nrow(main_pairs)) {
  for (i in seq_len(nrow(main_pairs))) {
    pair <- main_pairs[i, ]
    metric <- if (pair$score_type == "quantile_loss") "quantile loss" else "joint VaR/ES score"
    label <- sprintf("At %.1f%% confidence, copula-GARCH versus **%s** has mean %s difference **%s**",
                     100 * pair$confidence, pair$reference_model, metric,
                     format(pair$mean_loss_difference, digits = 4, scientific = TRUE))
    if (pair$status != "ok") {
      cat(sprintf("%s. Paired inference is unavailable: `%s`.\n\n", label, pair$status))
    } else {
      direction <- if (pair$ci_upper < 0) "The marginal interval lies below zero." else if (pair$ci_lower > 0) {
        "The marginal interval lies above zero."
      } else "The interval crosses zero, so its direction is unresolved."
      cat(sprintf("%s, with a %.1f%% interval [%s, %s] and Holm-adjusted p-value %s. %s\n\n",
                  label, 100 * pair$confidence_interval,
                  format(pair$ci_lower, digits = 4, scientific = TRUE),
                  format(pair$ci_upper, digits = 4, scientific = TRUE),
                  format(pair$p_value_holm, digits = 4), direction))
    }
  }
} else {
  cat("The main copula-GARCH model is absent from this run's paired comparisons; this run does not evaluate its incremental predictive value.\n")
}
```

```{r}
show_columns(predictive, c("model", "reference_model", "confidence", "score_type",
                          "common_sample_fraction", "block_length", "bootstrap_replications",
                          "p_value_holm", "status"), digits = 5)
```

Each confidence-level and score comparison uses a common date set across all models.
Inference is reported when that sample is sufficiently large and contiguous on the
forecast calendar. The mean difference and availability remain informative when
these conditions are unmet. Per-date paired scores in `predictive_scores.csv`
allow individual periods and loss contributions to be examined.

# Monte Carlo precision

```{r}
convergence <- read_output("monte_carlo_convergence_summary", required = FALSE)
if (!is.null(convergence)) {
  show_columns(convergence, c("n_simulations", "confidence", "repetitions", "mean_var", "sd_var",
                              "mean_es", "sd_es"), digits = 6)
} else {
  knitr::kable(data.frame(Status = "The convergence experiment was not enabled for this run."))
}
```

```{r}
show_figure("monte_carlo_convergence.png")
```

The repeated-seed experiment measures Monte Carlo variation conditional on the fitted
model. Its convergence pattern helps assess whether the simulation budget is adequate
for the reported risk estimates. Parameter estimation, model selection and structural
change contribute additional uncertainty beyond this experiment. Reproduction uses
the recorded draw count, seed and chunk size together.

# Computational cost

```{r}
show_columns(comparison, c("model", "confidence", "runtime_seconds", "model_failures"), digits = 3)
```

The timing table reports elapsed wall-clock time attributed to each model. The
fixed-marginal copula comparisons share marginal estimation, and their attributed
times include that common cost. Fits and simulations can also support several
confidence levels. Pipeline runtime is therefore distinct from the sum of these rows.

# Interpretation and scope

The principal empirical question is whether multivariate modelling delivers enough
predictive gain to justify its additional estimation and simulation cost. The paired
means measure the magnitude of that gain relative to portfolio GARCH-t. Holding
marginal forecasts fixed then identifies the role of dependence assumptions in the
observed ordering of models. An improvement in average loss and uncertainty about its
persistence can coexist; intervals spanning zero warrant further evaluation across
samples and market regimes.

The experiment covers a bivariate index portfolio with fixed weights. Its
local-currency convention leaves FX risk, hedging and trading costs outside the
portfolio definition. Sparse extreme losses constrain tail validation, while estimated
marginal parameters and dependence structures add uncertainty beyond Monte Carlo
variation. These boundaries guide the interpretation of the measured gains and the
design of further comparisons.

# Reproducibility

```{r}
keys <- c("run_id", "profile", "git_commit", "source_dirty", "implementation_hash", "config_hash",
          "input_hash", "aligned_hash", "retrieved_at_utc", "r_version", "rng_kind",
          "simulation_seed", "current_draws", "rolling_draws", "chunk_size")
knitr::kable(provenance[match(keys, provenance$key, nomatch = 0L), ], row.names = FALSE)
```

The experiment records its source revision, configuration, software environment and
input identities. The run archive contains the price snapshot, configuration and
session information; the output manifest connects these inputs to the generated tables
and figures. Package versions are listed in `run_provenance.csv`. Repeating an
experiment requires both its software environment and its historical input snapshot,
since later provider downloads can revise the underlying data.

# References

::: {#refs}
:::
